#Reproducible Research: Peer Assessment 1

####Courtesy  
sefakilic@github  
Amit Upadhyay@RPubs  
http://rstudio-pubs-static.s3.amazonaws.com/19424_092bcbaab3ee4aa6b470e9a2c3c66ee3.html

###Loading and preprocessing the data

```{r}
library(knitr)
library(data.table)
library(ggplot2)

opts_chunk$set(echo = TRUE, results = 'hold')
        
act <- read.csv('activity.csv', header = TRUE, sep = ",",
                  colClasses=c("numeric", "character", "numeric"))
act$date <- as.Date(act$date, format = "%Y-%m-%d")
act$interval <- as.factor(act$interval)
str(act)
```

###What is mean total number of steps taken per day?

Use aggregate function instead of tapply.   
Add column names for easy identification.  
And finally display a few records in this table.  

```{r}
library(ggplot2)
steps_per_day <- aggregate(steps ~ date, act, sum)
colnames(steps_per_day) <- c("date","steps")
head(steps_per_day)
```

Use ggplot with parameters as below.

```{r}
ggplot(steps_per_day, aes(x = steps)) + 
       geom_histogram(fill = "light green", binwidth = 1000) + 
        labs(title="Histogram of Steps Taken per Day", 
             x = "Number of Steps per Day", 
y = "Number of times in a day(Count)") + theme_bw()
```

```{r}
steps_mean   <- mean(steps_per_day$steps, na.rm=TRUE)
steps_mean
steps_median <- median(steps_per_day$steps, na.rm=TRUE)
steps_median
```

###What is the average daily activity pattern?

Use aggregate fnction to calculate the mean for every interval. This would also be
useful for imputing values later.

```{r}
library(ggplot2)

steps_per_interval <- aggregate(act$steps, 
                                by = list(interval = act$interval),
                                FUN=mean, na.rm=TRUE)
####convert to integers
####this helps in plotting
steps_per_interval$interval <- 
        as.integer(levels(steps_per_interval$interval)[steps_per_interval$interval])
colnames(steps_per_interval) <- c("interval", "steps")

ggplot(steps_per_interval, aes(x=interval, y=steps)) +   
        geom_line(color="orange", size=1) +  
        labs(title="Average Daily Activity Pattern", x="Interval", y="Number of steps") +  
        theme_bw()
```

On average across all the days in the dataset, the 5-minute interval contains the maximum number of steps?

'which' function is used for this purpose.

```{r}
max_interval <- steps_per_interval[which.max(  
        steps_per_interval$steps),]
max_interval
```

###Imputing missing values

There are many days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data. To overcome this we will have to impute values. In this exercise we will impute
with mean values for that specific interval.

#### How many missing values are in this dataset?

```{r}
missing_vals <- sum(is.na(act$steps))
missing_vals
```

#####All of the missing values are filled in with mean value for that 5-minute interval.

To populate missing values, we choose to replace them with the mean value at the same interval across days. In most of the cases the median is a better centrality measure than mean, but in this we will use mean as there is not a significant difference
between them.  

We will create a function na_fill(data, pervalue) which takes two arguments.  
1. Data frame - act  
2. Steps per interval in the act dataframe  

```{r}
na_fill <- function(data, pervalue) {
        na_index <- which(is.na(data$steps))
        na_replace <- unlist(lapply(na_index, FUN=function(idx){
                interval = data[idx,]$interval
                pervalue[pervalue$interval == interval,]$steps
        }))
        fill_steps <- data$steps
        fill_steps[na_index] <- na_replace
        fill_steps
}

act_fill <- data.frame(  
        steps = na_fill(act, steps_per_interval),  
        date = act$date,  
        interval = act$interval)
str(act_fill)

#We check if there are any missing values remaining. Expected count is 0.

sum(is.na(act_fill$steps))

fill_steps_per_day <- aggregate(steps ~ date, act_fill, sum)
colnames(fill_steps_per_day) <- c("date","steps")

##plotting the histogram
ggplot(fill_steps_per_day, aes(x = steps)) + 
       geom_histogram(fill = "blue", binwidth = 1000) + 
        labs(title="Histogram of Steps Taken per Day", 
             x = "Number of Steps per Day", y = "Number of times in a day(Count)") + theme_bw()

```

Mean and median values are higher after imputing missing data. The reason is that in the original data, there are some days with steps values NA for all intervals. The total number of steps taken in such days are set to 0s by default. However, after replacing missing steps values with the mean steps of associated interval value, these 0 values are removed from the histogram of total number of steps taken each day.

```{r}
steps_mean_fill   <- mean(fill_steps_per_day$steps, na.rm=TRUE)
steps_mean_fill
steps_median_fill <- median(fill_steps_per_day$steps, na.rm=TRUE)
steps_median_fill
```

###Are there differences in activity patterns between weekdays and weekends?

We do this comparison with the table with filled-in missing values.  
1. Augment the table with a column that indicates the day of the week    
2. Subset the table into two parts - weekends (Saturday and Sunday) and weekdays (Monday through Friday).  
3. Tabulate the average steps per interval for each data set.  
4. Plot the two data sets side by side for comparison.  

First, we will find the day of the week for each measurement in the dataset. In this part, we use the dataset with the filled-in values.

```{r}
weekdays_steps <- function(data) {
    weekdays_steps <- aggregate(data$steps, by=list(interval = data$interval),
                          FUN=mean, na.rm=TRUE)
    # convert to integers for plotting. Similar to one of the earlier steps.
    weekdays_steps$interval <- 
            as.integer(levels(weekdays_steps$interval)[weekdays_steps$interval])
    colnames(weekdays_steps) <- c("interval", "steps")
    weekdays_steps
}

data_by_weekdays <- function(data) {
    data$weekday <- 
            as.factor(weekdays(data$date)) # weekdays
    weekend_data <- subset(data, weekday %in% c("Saturday","Sunday"))
    weekday_data <- subset(data, !weekday %in% c("Saturday","Sunday"))

    weekend_steps <- weekdays_steps(weekend_data)
    weekday_steps <- weekdays_steps(weekday_data)

    weekend_steps$dayofweek <- rep("weekend", nrow(weekend_steps))
    weekday_steps$dayofweek <- rep("weekday", nrow(weekday_steps))

    data_by_weekdays <- rbind(weekend_steps, weekday_steps)
    data_by_weekdays$dayofweek <- as.factor(data_by_weekdays$dayofweek)
    data_by_weekdays
}

data_weekdays <- data_by_weekdays(act_fill)
```

Below we see the panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends:

```{r}
ggplot(data_weekdays, aes(x=interval, y=steps)) + 
        geom_line(color="violet") + 
        facet_wrap(~ dayofweek, nrow=2, ncol=1) +
        labs(x="Interval", y="Number of steps") +
        theme_bw()
```

We can see at the graph above that activity on the weekday has the greatest peak from all steps intervals. We can also see that weekends activities has more peaks than a weekday. This could be because during work hours there are limited activity but whereas in the weekend there could be burst of activities at various time intervals.